---
title: "CLY23 Phenotype Analysis: PCA and Heritability"
author: "Analysis Report"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    code_download: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 7
)

# Set working directory
knitr::opts_knit$set(root.dir = "/Users/fvrodriguez/Desktop/inv4m/heritability")
```

## Load Libraries

**Purpose**: Load required packages for data manipulation, visualization, and multivariate analysis.

**Approach**: Load tidyverse ecosystem and PCA visualization packages.

**Expected outcome**: All dependencies available for downstream analysis.

```{r load_libraries}
library(dplyr)
library(ggplot2)
library(lubridate)
library(MASS)
library(FactoMineR)
library(factoextra)
library(ggbeeswarm)
library(ggpubr)
```

## Import LI-COR Data

**Purpose**: Import photosynthesis measurements from local CSV.

**Approach**: Read PhiPS2 data from exported CSV, parse sample IDs to extract field coordinates.

**Expected outcome**: Clean dataframe with CLY23_D4 plot IDs and PhiPS2 values.

```{r import_licor}
# Read from local CSV (exported from Google Sheets)
licor_raw <- read.csv("licor_data.csv", skip = 1)

# Parse sample IDs
licor <- stringr::str_split(licor_raw$BZea, "_", simplify = TRUE) %>% 
  as.data.frame()
colnames(licor) <- c("field_year", "section", "CLY23_D4")
licor$CLY23_D4 <- as.numeric(licor$CLY23_D4)
licor$PhiPS2 <- as.numeric(licor_raw$PhiPS2)

# Diagnostics
{
cat("LI-COR records:", nrow(licor), "\n")
cat("PhiPS2 range:", range(licor$PhiPS2, na.rm = TRUE), "\n")
}
```

## Import and Process CLY23 Field Data

**Purpose**: Import main phenotype data and calculate derived traits.

**Approach**: Read field book CSV, calculate flowering time metrics (DTS, DTA, ASI), parse donor identifiers, merge with LI-COR data.

**Expected outcome**: Complete CLY23 dataframe with all phenotypes and donor metadata.

```{r import_cly23}
planting_date <- mdy("4/4/2023")

CLY23 <- read.csv("CLY23_D4_FieldBook.csv") %>%
  rename(silking = "DTS", anthesis = "DTA", pedigree = "Female_genotype") %>%
  mutate(
    DTS = (mdy(silking) - planting_date) %>% as.integer(),
    DTA = (mdy(anthesis) - planting_date) %>% as.integer(),
    dSPAD = SPAD2 - SPAD1,
    donor_pop = gsub("-.*", "", pedigree, perl = TRUE),
    donor_line = gsub("_P.*?$", "", pedigree, perl = TRUE)
  ) %>%
  left_join(licor %>% dplyr::select(CLY23_D4, PhiPS2))

# Clean donor_line strings
CLY23$donor_line <- gsub("^.*?-", "", CLY23$donor_line, perl = TRUE)
CLY23$donor_line <- gsub("-+", "_", CLY23$donor_line, perl = TRUE)
CLY23$donor_line <- gsub("\\(.*", "", CLY23$donor_line, perl = TRUE)
CLY23$donor_line <- gsub("-", "_", CLY23$donor_line, perl = TRUE)
CLY23$donor_line <- gsub(" +$", "", CLY23$donor_line, perl = TRUE)

# Calculate derived traits
CLY23$ASI <- CLY23$DTS - CLY23$DTA
CLY23$LL <- CLY23$SL + CLY23$BL
CLY23$StPu <- as.factor(CLY23$StPu) %>% as.numeric() - 1
CLY23$StPi <- as.factor(CLY23$StPi) %>% as.numeric() - 1
CLY23$donor_line <- factor(CLY23$donor_line)

# Diagnostics
{
cat("CLY23 records:", nrow(CLY23), "\n")
cat("Donor populations:", length(unique(CLY23$donor_pop)), "\n")
cat("Donor lines:", nlevels(CLY23$donor_line), "\n")
cat("Available traits:", paste(colnames(CLY23), collapse = ", "), "\n")
}
```

## Define Trait Categories

**Purpose**: Organize phenotypic variables into functional categories for analysis and visualization.

**Approach**: Create vectors of trait names and corresponding category labels.

**Expected outcome**: Trait and category vectors for downstream analysis.

```{r define_traits}
vars <- c("DTA", "DTS", "ASI", "PH", "EH", "EN", "BW", "BL", "SL", "LL", 
          "NBR", "SPAD1", "SPAD2", "dSPAD", "PhiPS2", "StPu", "StPi")

var_type <- c("FT", "FT", "FT", "Plant Size", "Ear", "Ear", "Plant Size", 
              "Plant Size", "Plant Size", "Plant Size", "Plant Size", 
              "Photosynthesis", "Photosynthesis", "Photosynthesis", "Photosynthesis", 
              "Highland", "Highland")

var_pal <- c("darkgoldenrod1", "darkred", "darkmagenta", "darkgreen", "burlywood")

# Diagnostics
{
cat("Traits defined:", length(vars), "\n")
cat("Categories:", paste(unique(var_type), collapse = ", "), "\n")
}
```

## Prepare PCA Data

**Purpose**: Aggregate phenotype data to genotype means for PCA.

**Approach**: Filter out controls, group by pedigree, calculate trait means across replicates.

**Expected outcome**: Genotype-level means suitable for multivariate analysis.

```{r prep_pca}
# Genotype categories for plotting
cats <- CLY23 %>%
  dplyr::filter(donor_pop != "Purple Check") %>%
  droplevels() %>%
  dplyr::select(donor_pop, donor_line, pedigree, Rep) %>%
  group_by(donor_pop, donor_line, pedigree) %>%
  summarize(n = n(), .groups = "drop")

# Genotype means for PCA
to_pca <- CLY23 %>%
  dplyr::select(donor_pop, donor_line, pedigree, Rep, all_of(vars)) %>%
  droplevels() %>%
  dplyr::filter(donor_pop != "Purple Check") %>%
  group_by(pedigree) %>%
  summarise(across(all_of(vars), ~ mean(.x, na.rm = TRUE)), .groups = "drop")

# Diagnostics
{
cat("Genotypes for PCA:", nrow(to_pca), "\n")
cat("Traits with complete data:", sum(colSums(is.na(to_pca[,-1])) == 0), "/", length(vars), "\n")
}
```

## Principal Component Analysis

**Purpose**: Reduce dimensionality and identify major axes of phenotypic variation.

**Approach**: Run PCA on genotype means, visualize individual and variable contributions.

**Expected outcome**: PCA plots showing genotype clustering by donor population and trait loadings.

```{r run_pca}
pca <- PCA(to_pca[, -1], graph = FALSE)

# Diagnostics
{
cat("=== PCA Variance Explained ===\n")
cat("PC1:", round(pca$eig[1, 2], 1), "%\n")
cat("PC2:", round(pca$eig[2, 2], 1), "%\n")
cat("Cumulative (PC1-2):", round(pca$eig[2, 3], 1), "%\n")
}
```

```{r pca_individuals, fig.cap="PCA of genotypes colored by donor population"}
fviz_pca_ind(
  pca, 
  col.ind = cats$donor_pop, 
  geom = "point", 
  pointsize = 2,
  addEllipses = TRUE
) +
  guides(
    col = guide_legend(title = "Donor", override.aes = aes(label = "")),
    shape = "none", 
    fill = "none"
  ) +
  theme_classic2(base_size = 15)
```

```{r pca_variables, fig.cap="PCA variable loadings by trait category"}
fviz_pca_var(pca, col.var = as.factor(var_type)) +
  scale_color_manual(values = var_pal) +
  guides(
    col = guide_legend(title = "Type", override.aes = aes(label = ""))
  ) +
  theme_classic2(base_size = 15)
```

## Trait Correlations

**Purpose**: Examine pairwise correlations among phenotypic traits.

**Approach**: Compute correlation matrix using pairwise complete observations.
  
**Expected outcome**: Correlation matrix revealing trait relationships.

```{r trait_correlations}
cor_matrix <- cor(to_pca[, -1], use = "pairwise.complete.obs")
round(cor_matrix, 2)
```

## Broad-Sense Heritability by Trait

**Purpose**: Estimate H² for each phenotypic trait.

**Approach**: For each trait, subset to genotypes with all 3 replicates, fit one-way ANOVA, partition variance.

**Expected outcome**: Heritability estimates for all measured traits.

```{r heritability_calculation}
her <- data.frame()

for (trait in vars) {
  # Remove NAs for this trait
  without_NA <- CLY23[!is.na(CLY23[, trait]), ]
  
  # Keep only genotypes with all 3 reps
  with_reps <- as.data.frame(table(line = without_NA$pedigree)) %>%
    filter(Freq == 3) %>%
    pull(line)
  
  to_H2 <- without_NA %>%
    filter(pedigree %in% with_reps)
  
  # Fit ANOVA model
  formula <- as.formula(paste0(trait, " ~ pedigree"))
  mod <- lm(data = to_H2, formula)
  mod_aov <- anova(mod)
  
  # Calculate H2
  her <- rbind(her, data.frame(
    trait = trait,
    H2 = mod_aov$`Sum Sq`[1] / sum(mod_aov$`Sum Sq`),
    n_geno = length(with_reps)
  ))
}

her$type <- var_type

# Diagnostics
{
cat("=== Heritability Estimates ===\n")
print(her %>% arrange(desc(H2)))
}
```

```{r heritability_plot, fig.cap="Broad-sense heritability by trait"}
her %>%
  mutate(trait = forcats::fct_reorder(trait, H2, .desc = TRUE)) %>%
  ggplot(aes(x = trait, y = H2, fill = type)) +
  geom_col() +
  ylab("Broad-Sense Heritability (H²)") +
  xlab("Trait") +
  scale_fill_manual(values = var_pal) +
  theme_classic2(base_size = 15) +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "top"
  )
```

## Import CLY22 Brace Root Data

**Purpose**: Import previous year's brace root data for cross-year comparison.

**Approach**: Read CLY22 field book, filter to J2Teo project, parse donor identifiers.

**Expected outcome**: Clean CLY22 dataframe for GxE analysis.

```{r import_cly22}
CLY22 <- read.csv("CLY-22-RRA-Field-Book - B6Field.csv") %>%
  dplyr::filter(Project == "J2Teo") %>%
  dplyr::mutate(
    donor_pop = gsub("-.*", "", Description, perl = TRUE),
    donor_line = gsub("_P.*?$", "", Description, perl = TRUE)
  )

CLY22$brace_root <- as.numeric(CLY22$brace_root)

# Clean donor_line strings
CLY22$donor_line <- gsub("^.*?-", "", CLY22$donor_line, perl = TRUE)
CLY22$donor_line <- gsub("-+", "_", CLY22$donor_line, perl = TRUE)
CLY22$donor_line <- gsub("\\(.*", "", CLY22$donor_line, perl = TRUE)
CLY22$donor_line <- gsub("-", "_", CLY22$donor_line, perl = TRUE)
CLY22$donor_line <- gsub(" +$", "", CLY22$donor_line, perl = TRUE)

# Diagnostics
{
cat("CLY22 J2Teo records:", nrow(CLY22), "\n")
cat("Brace root range:", range(CLY22$brace_root, na.rm = TRUE), "\n")
}
```

## Combine Brace Root Data Across Years

**Purpose**: Merge brace root measurements from CLY22 and CLY23 for GxE analysis.

**Approach**: Standardize column names, stack datasets with field year indicator.

**Expected outcome**: Combined dataframe for cross-environment analysis.

```{r combine_br_data}
BR <- rbind(
  CLY22 %>%
    dplyr::select(donor_line, donor_pop, NBR = brace_root) %>%
    dplyr::filter(donor_pop != "Purple Check") %>%
    mutate(field = "CLY22"),
  
  CLY23 %>%
    dplyr::filter(Rep == 3) %>%
    dplyr::select(donor_line, donor_pop, NBR) %>%
    dplyr::filter(donor_pop != "Purple Check") %>%
    mutate(field = "CLY23")
)

field_mean <- BR %>%
  group_by(field) %>%
  summarise(br_mean = mean(NBR, na.rm = TRUE), .groups = "drop")

# Diagnostics
{
cat("=== Combined Brace Root Data ===\n")
print(table(BR$field, BR$donor_pop))
cat("\nField means:\n")
print(field_mean)
}
```

## Brace Root Variation by Population and Year

**Purpose**: Visualize brace root number variation across donor populations and field years.

**Approach**: Faceted beeswarm plots with population means and field-level reference lines.

**Expected outcome**: Visual comparison of NBR distributions across populations and environments.

```{r br_population_plot, fig.cap="Brace root number by donor population and field year", fig.height=9}
BR %>%
  ggplot(aes(x = donor_pop, y = NBR, group = donor_pop, col = donor_pop)) +
  geom_quasirandom() +
  geom_hline(
    data = field_mean,
    aes(yintercept = br_mean),
    colour = "black", linetype = 2
  ) +
  stat_summary(
    fun.data = mean_cl_normal,
    geom = "pointrange",
    col = "black"
  ) +
  facet_wrap(~field, ncol = 1) +
  scale_y_continuous(breaks = 0:6) +
  xlab("Donor Population") +
  ylab("Number of Brace Roots") +
  theme_classic2(base_size = 15) +
  theme(
    legend.position = "none",
    strip.background = element_rect(colour = "white", fill = "white"),
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)
  )
```

## Brace Root Heritability by Year

**Purpose**: Estimate heritability of brace root number within each field year.

**Approach**: Fit one-way ANOVA (NBR ~ donor_line) separately for each year.

**Expected outcome**: Year-specific H² estimates for brace root number.

```{r br_heritability_by_year}
# CLY22 heritability
br22_aov <- lm(formula = NBR ~ donor_line, data = BR[BR$field == "CLY22", ])
s22 <- summary(br22_aov)

# CLY23 heritability
br23_aov <- lm(formula = NBR ~ donor_line, data = BR[BR$field == "CLY23", ])
s23 <- summary(br23_aov)

# Diagnostics
{
cat("=== Brace Root H² by Year ===\n")
cat("CLY22 H² (R²):", round(s22$r.squared, 3), "\n")
cat("CLY23 H² (R²):", round(s23$r.squared, 3), "\n")
}
```

## Genotype × Environment Analysis

**Purpose**: Partition variance into genotype, environment, and GxE components for brace root.

**Approach**: Fit two-way ANOVA with interaction (NBR ~ donor_line * field), extract variance components.

**Expected outcome**: Proportion of variance explained by genotype, field year, and their interaction.

```{r gxe_analysis}
# Full GxE model
br_gxe_mod <- lm(formula = NBR ~ donor_line * field, data = BR)
gxe_aov <- anova(br_gxe_mod)

# Variance components
var_components <- data.frame(
  source = c("Genotype", "Field", "G×E", "Residual"),
  sum_sq = gxe_aov$`Sum Sq`,
  prop_var = gxe_aov$`Sum Sq` / sum(gxe_aov$`Sum Sq`)
)

# Summary table
gxe_summary <- data.frame(
  field = c("CLY22", "CLY23", "Combined"),
  H2_fam = c(
    s22$r.squared, 
    s23$r.squared, 
    gxe_aov$`Sum Sq`[1] / sum(gxe_aov$`Sum Sq`)
  ),
  field_year = c(NA, NA, gxe_aov$`Sum Sq`[2] / sum(gxe_aov$`Sum Sq`)),
  GxFY = c(NA, NA, gxe_aov$`Sum Sq`[3] / sum(gxe_aov$`Sum Sq`))
)

# Diagnostics
{
cat("=== Variance Components ===\n")
print(var_components)
cat("\n=== GxE Summary ===\n")
print(gxe_summary)
}
```

## Population-Level GxE Analysis

**Purpose**: Estimate heritability and GxE within each donor population.

**Approach**: Split data by population, fit GxE models where both years are represented.

**Expected outcome**: Population-specific variance partitioning.

```{r population_gxe}
BR$field_donor <- interaction(factor(BR$field), factor(BR$donor_pop))

# Diagnostic: show donor_line counts per field × donor_pop combination
line_counts <- BR %>%
  group_by(field, donor_pop) %>%
  summarise(
    n_obs = n(),
    n_lines = n_distinct(donor_line),
    .groups = "drop"
  ) %>%
  arrange(n_lines)

# Diagnostics
{
cat("=== Donor Lines per Field × Population ===\n")
cat("(Combinations with n_lines < 2 will be skipped)\n\n")
print(as.data.frame(line_counts))
cat("\nSkipped combinations:", sum(line_counts$n_lines < 2), "\n")
}

# H2 by population within each field
br_list_field <- split(BR, f = BR$field_donor)

H2_pop <- lapply(names(br_list_field), FUN = function(x) {
  df <- br_list_field[[x]]
  n_lines <- length(unique(df$donor_line))
  if (nrow(df) < 3 || n_lines < 2) return(NULL)
  
  df$donor_line <- factor(df$donor_line)
  lm_aov <- lm(data = df, NBR ~ donor_line) %>% anova()
  
  data.frame(
    field = df$field[1],
    donor_pop = df$donor_pop[1],
    H2_fam = lm_aov$`Sum Sq`[1] / sum(lm_aov$`Sum Sq`)
  )
}) %>% 
  dplyr::bind_rows()

# GxE by population (across fields)
br_list_pop <- split(BR, f = BR$donor_pop)

# Diagnostic: show which populations have data in both fields
pop_field_summary <- BR %>%
  group_by(donor_pop) %>%
  summarise(
    n_fields = n_distinct(field),
    n_lines = n_distinct(donor_line),
    .groups = "drop"
  )

# Diagnostics
{
cat("\n=== Population Summary for GxE Analysis ===\n")
cat("(Populations with n_fields < 2 or n_lines < 2 will be skipped)\n\n")
print(as.data.frame(pop_field_summary))
}

H2_pop_gxe <- lapply(names(br_list_pop), FUN = function(x) {
  df <- br_list_pop[[x]]
  field_count <- table(df$donor_line, df$field) %>% as.data.frame()
  n_field <- length(unique(field_count[, 2]))
  n_lines <- length(unique(df$donor_line))
  
  if (n_field < 2 || n_lines < 2) return(NULL)
  
  df$donor_line <- factor(df$donor_line)
  lm_aov <- lm(data = df, NBR ~ donor_line * field) %>% anova()
  
  data.frame(
    donor_pop = df$donor_pop[1],
    H2_fam = lm_aov$`Sum Sq`[1] / sum(lm_aov$`Sum Sq`),
    GxF = lm_aov$`Sum Sq`[3] / sum(lm_aov$`Sum Sq`),
    field = lm_aov$`Sum Sq`[2] / sum(lm_aov$`Sum Sq`),
    residuals = lm_aov$`Sum Sq`[4] / sum(lm_aov$`Sum Sq`)
  )
}) %>% 
  dplyr::bind_rows()

# Diagnostics
{
cat("=== H² by Population and Field ===\n")
print(H2_pop)
cat("\n=== Population-Level GxE ===\n")
print(H2_pop_gxe)
}
```

## Session Info

```{r session_info}
sessionInfo()
```
